Refinement of Thermostated Molecular Dynamics via Backward Error Analysis

### 2018-06-09 rigid_bodies.tex
Revision of the Appendix

For two non-commutative Liouville operators $\Liu A$ and $\Liu B$ and a real constant $\timestep$, the symmetric Baker-Campbell-Hausdorff (BCH) formula can be expressed as \cite{Hairer_2006}
\begin{equation}
\label{eq:symmetric BCH}
\begin{split}
&e^{\frac{\timestep}{2} \Liu B} e^{\timestep \Liu A} e^{\frac{\timestep}{2} \Liu B} = \\
&= e^{\timestep (\Liu A + \Liu B) + \frac{\timestep^3}{24} \left[2 \Liu A + \Liu B,[\Liu A,\Liu B]\right] + \mathcal{O}(\timestep^5)},
\end{split}
\end{equation}
where $[X,Y]$ is the commutator of operators $X$ and $Y$, defined as $[X,Y] = XY - YX$.
The remaining terms of the infinite series involve ever-deepening, nested commutators \cite{Hairer_2006}.

Let us now represent the operators $\Liu A$ and $\Liu B$ using Poisson brackets involving Hamiltonians $\Ham A$ and $\Ham B$, that is, $\Liu A = \{\circ,\Ham A\}$ and $\Liu B = \{\circ,\Ham B\}$.
By employing the properties of anti-commutativity and Jacobi identity \cite{Hairer_2006}, we can deduce that
\begin{align*}
[\Liu A,\Liu B] &= \{\{\circ,\Ham B\},\Ham A\} - \{\{\circ,\Ham A\},\Ham B\} = \\
&= -\{\Ham A,\{\circ,\Ham B\}\} - \{\Ham B,\{\Ham A,\circ\}\} = \\
&= \{\circ,\{\Ham B,\Ham A\}\} = \\
&= \{\circ,{\Liu A} {\Ham B}\}.
\end{align*}

This means that the commutator $[\Liu A,\Liu B]$ is, in fact, a new Liouville operator $\Liu C$ whose associated Hamiltonian is $\Ham C = {\Liu A}{\Ham B}$.
By applying this procedure recursively, we can rewrite the right-hand side of Eq.~\eqref{eq:symmetric BCH} as $e^{\timestep \Liu S}$, where $\Liu S$ is the Liouville operator whose associated Hamiltonian is
\begin{equation*}
\label{eq:general shadow hamiltonian}
\refined{\Ham{}} = \Ham A + \Ham B + \frac{\timestep^2}{24} (2 \Liu A + \Liu B){\Liu A}{\Ham B} + \mathcal{O}(\timestep^4).
\end{equation*}

Therefore, a splitting method meant to approximately reproduce the dynamics of a system with Hamiltonian $\Ham{} = \Ham A + \Ham B$ will, in fact, reproduce exactly (putting round-off issues aside) the dynamics of another system with Hamiltonian $\refined{\Ham{}}$ as above.
In the literature, $\refined{\Ham{}}$ is usually referred to as a shadow Hamiltonian.

In the development that follows, we consider a single rigid body and, in the end, we generalize the result for a system of rigid bodies.

For the splitting introduced in Section~\ref{sec:nve}, $\Ham A = K(\vt p, \vt q, \vt \pi)$ and $\Ham B = U(\vt r, \vt q)$ with the corresponding Liouviller operators $\Liu{A} = \tr{K_{\vt p}}\diff{}{\vt r} + \tr{K_{\vt \pi}}\diff{}{\vt q} - \tr{K_{\vt q}}\diff{}{\vt \pi}$ and $\Liu{B} = -\tr{U_{\vt r}}\diff{}{\vt p} - \tr{U_{\vt q}}\diff{}{\vt \pi}$, with $K_{x}$ and $U_{x}$ being the gradient of the kinetic and potential energies, respectively.
We first obtain $\Liu A \Ham B = \tr{K_{\vt p}} U_{\vt r} + \tr{K_{\vt \pi}} U_{\vt q}$, and we find that
\begin{align*}
\Liu A \Liu A \Ham B &= \tr{K_{\vt p}} U_{\vt r \vt r} K_{\vt p}
+ 2 \tr{K_{\vt \pi}} U_{\vt r \vt q} K_{\vt p}
+ \tr{K_{\vt \pi}} U_{\vt q \vt q} K_{\vt \pi} \\
&+ \tr{K_{\vt \pi}} K_{\vt \pi \vt q} U_{\vt q}
- \tr{K_{\vt q}} K_{\vt \pi \vt \pi} U_{\vt q},
\end{align*}
where the fact that $\tr{U_{\vt q \vt r}} = U_{\vt r \vt q}$ has been exploited.
As reported in Ref.~\citenum{Silveira_2017}, the identity $\tr{\mt B}(\vt q) {\vt \pi} = -\tr{\mt B}(\vt \pi) {\vt q}$ is useful for performing differentiations.
The first-order derivatives of $K$ with respect to $\vt p$, $\vt q$, and $\vt \pi$ are, respectively,
\begin{align*}
&K_{\vt p} = {\mt M}^{-1} {\vt p}, \\
&K_{\vt q} = -\frac{1}{2} {\mt B}(\vt \pi) {\vt \omega}, \; \text{and} \\
&K_{\vt \pi} = \frac{1}{2} {\mt B}(\vt q) {\vt \omega},
\end{align*}
where $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) \vt \pi = -\frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt \pi) \vt q$.\cite{Silveira_2017} Hence,
\begin{gather*}
\tr{K_{\vt p}} U_{\vt r \vt r} K_{\vt p} = \tr{\vt p} {\mt M}^{-1} U_{\vt r \vt r} {\mt M}^{-1} {\vt p}, \\
\tr{K_{\vt \pi}} U_{\vt r \vt q} K_{\vt p} = \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt q) U_{\vt r \vt q} {\mt M}^{-1} {\vt p}, \\
\tr{K_{\vt \pi}} U_{\vt q \vt q} K_{\vt \pi} = \frac{1}{4} \tr{\vt \omega} \tr{\mt B}(\vt q) U_{\vt q \vt q} {\mt B}(\vt q) \vt \omega.
\end{gather*}

@@

Pre-multiplication by $\tr{K_{\vt q}}$ will introduce a product $\tr{\mt B}(\vt \pi) {\mt B}(\vt q)$, which has been shown in Ref.~\citenum{Silveira_2017} to be equal to ${\mt S}\left( \tr{\mt B}(\vt \pi) {\vt q} \right) = -2 {\mt S}({\mt I} {\vt \omega})$, where the operator ${\mt S}(\cdot)$ builds a skew-symmetric matrix from the entries of a vector.
In addition, post-multiplication by $U_{\vt q}$ will introduce a product $\tr{\mt B}(\vt q) U_{\vt q}$, which is equal to $-2 {\vt \tau}^b$, where ${\vt \tau}^b$ is the body-fixed frame representation of the torque.
Then, the fact that ${\mt S}(\vt x) {\vt y} = -{\mt S}(\vt y) {\vt x}$ leads to
\begin{equation*}
\tr{K_{\vt q}} K_{\vt \pi \vt \pi} U_{\vt q} = \frac{1}{2} \tr{\vt \omega} {\mt S}({\mt I}^{-1} {\vt \tau}^b) {\mt I} {\vt \omega} = 0,
\end{equation*}
which is a null term as it corresponds to the quadratic form of a skew-symmetric matrix.
Another helpful identity taken from Appendix B of Ref.~\citenum{Silveira_2017} is ${\mt B}(\vt q){\vt \omega} = ( \sum_{k=1}^3 \omega_k \hat{\mt B}_k ) \vt q$, where each $\hat{\mt B}_k$ is a skew-symmetric permutation matrix (i.e.
$\tr{\hat{\mt B}}_k = -\hat{\mt B}_k$).
This allows us to obtain
\begin{align*}
K_{\vt \pi \vt q} &= \frac{1}{2} (\diff{\tr{\vt \omega}}{\vt q}) \tr{\mt B}(\vt q) + \frac{1}{2} \sum_{k=1}^3 \omega_k \tr{\hat{\mt B}}_k = \\
&= -\frac{1}{4} {\mt B}(\vt \pi) {\mt I}^{-1} \tr{\mt B}(\vt q) - \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k.
\end{align*}

Pre-multiplication by $\tr{K_{\vt \pi}}$ and post-multiplying $U_{\vt q}$, followed by some algebraic transformations, ultimately lead to
\begin{equation*}
\tr{K_{\vt \pi}} K_{\vt \pi \vt q} U_{\vt q} = -\frac{1}{2} \tr{\vt \omega} \tr{\mt S}({\mt I}^{-1} {\vt \tau}^b) {\mt I} {\vt \omega} + \frac{1}{2} \tr{\vt \omega} {\mt S}({\vt \tau}^b){\vt \omega}.
\end{equation*}
Both terms in the right-hand side of the equation above are identically null. 
Now, we are able to evaluate $\Liu A \Liu A \Ham B $, which is
\begin{align*}
\Liu A \Liu A \Ham B = &\tr{\vt p} {\mt M}^{-1} U_{\vt r \vt r} {\mt M}^{-1} {\vt p} + \tr{\vt \omega} \tr{\mt B}(\vt q) U_{\vt r \vt q} {\mt M}^{-1} {\vt p}
+ \\
&\frac{1}{4} \tr{\vt \omega} \tr{\mt B}(\vt q) U_{\vt q \vt q} {\mt B}(\vt q) \vt \omega.
\end{align*}
The remaining term to obtain $H_s$ is simply given by 
\begin{align*}
\Liu B \Liu A \Ham B &= -\tr{U_{\vt r}} \tr{K_{\vt p \vt p}} U_{\vt r} - \tr{U_{\vt q}} \tr{K_{\vt \pi \vt \pi}} U_{\vt q} = \\
&= -\tr{\vt F} {\mt M}^{-1} {\vt F} - \tr{\vt \tau}_b {\mt I}^{-1} {\vt \tau}_b.
\end{align*}

As will become clear below, it is convenient to rewrite the total kinetic energy of the original system, Eq.~\eqref{eq:kinetic energy}, as a quadratic form
\begin{equation*}
K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} {\mt \Omega}(\vt q) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
\end{equation*}
where $\mt \Omega$ is a symmetric, block-diagonal matrix of size $7N \times 7N$ defined as
\begin{equation}
\label{eq:block-diagonal inverse mass tensor}
{\mt \Omega} = \left[\begin{array}{cc}
{\mt M}^{-1} & \mt 0 \\
\mt 0 & \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q)
\end{array}\right].
\end{equation}

We now proceed to show the final expression of $H_s$, which results from grouping the terms corresponding to forces and torques in the \textit{modified} potential energy $\tilde{U}$, given by
\begin{equation*}
\refined U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A} {\mt I}^{-1} {\mt A} {\vt \tau} \right).
\end{equation*}
while the remaining terms can be recast in a compact matrix, as follows
\begin{equation*}
\refined K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \refined{\mathbf \Omega}(\vt r, \vt q) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
\end{equation*}
where
\begin{equation*}
\tilde{\mt \Omega} = {\mt \Omega} + \frac{\timestep^2}{6} \left( {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega} \right)
\end{equation*}
and
\begin{equation*}
{\mt U}_{\vt x \vt x} = \left[\begin{array}{cc}
U_{\vt r \vt r} & U_{\vt r \vt q} \\
U_{\vt q \vt r} & U_{\vt q \vt q}
\end{array}\right].
\end{equation*}

Finally, we have that
\begin{equation*}
\Ham S = \frac{1}{2} \tr{\left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \tilde{\mt \Omega} \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right] + \tilde{U} + \mathcal{O}({\timestep}^4),
\end{equation*}

Although $\refined{\mathbf \Omega}$ is still symmetric, it is no longer a block-diagonal matrix like $\mt \Omega$, due to the inter-body interactions accounted for in $U(\vt r, \vt q)$ and to the dependency of forces and torques on both positions and orientations.
As a result, the modified kinetic energy cannot, in principle, be split into independent translational and rotational contributions.

>>>>

For two non-commutative Liouville operators $\Liu A$ and $\Liu B$ and a real constant $\timestep$, the symmetric Baker-Campbell-Hausdorff (BCH) formula can be expressed as \cite{Hairer_2006}
\begin{equation}
\label{eq:symmetric BCH}
\begin{split}
&e^{\frac{\timestep}{2} \Liu B} e^{\timestep \Liu A} e^{\frac{\timestep}{2} \Liu B} = \\
&= e^{\timestep (\Liu A + \Liu B) + \frac{\timestep^3}{24} \left[2 \Liu A + \Liu B,[\Liu A,\Liu B]\right] + \mathcal{O}(\timestep^5)},
\end{split}
\end{equation}
where $[X,Y] = XY - YX$ is the commutator of operators $X$ and $Y$.
The remaining terms of the infinite series involve ever-deepening nested commutators \cite{Hairer_2006}.

Let us now represent the operators $\Liu A$ and $\Liu B$ using Poisson brackets involving Hamiltonians $\Ham A$ and $\Ham B$, that is, $\Liu A = \{\circ,\Ham A\}$ and $\Liu B = \{\circ,\Ham B\}$.
By employing the property of anti-commutativity and the Jacobi identity \cite{Hairer_2006}, we can deduce that
\begin{align*}
[\Liu A,\Liu B] &= \{\{\circ,\Ham B\},\Ham A\} - \{\{\circ,\Ham A\},\Ham B\} = \\
&= -\{\Ham A,\{\circ,\Ham B\}\} - \{\Ham B,\{\Ham A,\circ\}\} = \\
&= \{\circ,\{\Ham B,\Ham A\}\} = \\
&= \{\circ,{\Liu A} {\Ham B}\}.
\end{align*}

This means that the commutator $[\Liu A,\Liu B]$ is, in fact, a new Liouville operator $\Liu C$ whose associated Hamiltonian is $\Ham C = {\Liu A}{\Ham B}$.
By applying this procedure recursively, we can rewrite the right-hand side of Eq.~\eqref{eq:symmetric BCH} as $e^{\timestep \Liu S}$, where $\Liu S$ is the Liouville operator whose associated Hamiltonian is
\begin{equation*}
\label{eq:general shadow hamiltonian}
\Ham{S} = \Ham A + \Ham B + \frac{\timestep^2}{24} (2 \Liu A + \Liu B){\Liu A}{\Ham B} + \mathcal{O}(\timestep^4).
\end{equation*}

Therefore, a splitting method meant to approximately reproduce the dynamics of a system with Hamiltonian $\Ham{} = \Ham A + \Ham B$ will, in fact, reproduce exactly (round-off issues aside) the dynamics of another system with Hamiltonian $\Ham{S}$ as above.
In the literature, $\Ham{S}$ is usually referred to as a shadow Hamiltonian.

We remark that a single rigid body is considered in the development that follows, but the final result can be readily generalized for a system with multiple bodies.

For the splitting introduced in Sec.~\ref{sec:nve}, $\Ham A = K(\vt p, \vt q, \vt \pi)$ and $\Ham B = U(\vt r, \vt q)$ with the corresponding Liouville operators $\Liu{A} = \tr{K_{\vt p}}\diff{}{\vt r} + \tr{K_{\vt \pi}}\diff{}{\vt q} - \tr{K_{\vt q}}\diff{}{\vt \pi}$ and $\Liu{B} = -\tr{U_{\vt r}}\diff{}{\vt p} - \tr{U_{\vt q}}\diff{}{\vt \pi}$.
Note that any gradient like $\diff{f}{\vt x}$ is now represented as $f_{\vt x}$ for the sake of simplicity.
We start by obtaining $\Liu A \Ham B = \tr{K_{\vt p}} U_{\vt r} + \tr{K_{\vt \pi}} U_{\vt q}$. Next, we deduce that
\begin{align*}
\Liu A \Liu A \Ham B &= \tr{K_{\vt p}} U_{\vt r \vt r} K_{\vt p}
+ 2 \tr{K_{\vt \pi}} U_{\vt r \vt q} K_{\vt p}
+ \tr{K_{\vt \pi}} U_{\vt q \vt q} K_{\vt \pi} \\
&+ \tr{K_{\vt \pi}} K_{\vt \pi \vt q} U_{\vt q}
- \tr{K_{\vt q}} K_{\vt \pi \vt \pi} U_{\vt q},
\end{align*}
where the fact that $\tr{U_{\vt q \vt r}} = U_{\vt r \vt q}$ has been used.
As reported in Ref.~\citenum{Silveira_2017}, the identity $\tr{\mt B}(\vt q) {\vt \pi} = -\tr{\mt B}(\vt \pi) {\vt q}$ is useful for performing differentiations.
The first-order derivatives of $K$ with respect to $\vt p$, $\vt q$, and $\vt \pi$ are, respectively,
\begin{align*}
&K_{\vt p} = {\mt M}^{-1} {\vt p}, \\
&K_{\vt q} = -\frac{1}{2} {\mt B}(\vt \pi) {\vt \omega}, \; \text{and} \\
&K_{\vt \pi} = \frac{1}{2} {\mt B}(\vt q) {\vt \omega},
\end{align*}
where $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) \vt \pi = -\frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt \pi) \vt q$.\cite{Silveira_2017} Hence,
\begin{gather*}
\tr{K_{\vt p}} U_{\vt r \vt r} K_{\vt p} = \tr{\vt p} {\mt M}^{-1} U_{\vt r \vt r} {\mt M}^{-1} {\vt p}, \\
\tr{K_{\vt \pi}} U_{\vt r \vt q} K_{\vt p} = \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt q) U_{\vt r \vt q} {\mt M}^{-1} {\vt p}, \\
\tr{K_{\vt \pi}} U_{\vt q \vt q} K_{\vt \pi} = \frac{1}{4} \tr{\vt \omega} \tr{\mt B}(\vt q) U_{\vt q \vt q} {\mt B}(\vt q) \vt \omega.
\end{gather*}

@@

Pre-multiplication by $\tr{K_{\vt q}}$ will introduce a product $\tr{\mt B}(\vt \pi) {\mt B}(\vt q)$.
This has been shown in Ref.~\citenum{Silveira_2017} to be equal to ${\mt S}\left( \tr{\mt B}(\vt \pi) {\vt q} \right) = -2 {\mt S}({\mt I} {\vt \omega})$, where the operator ${\mt S}(\cdot)$ builds a skew-symmetric matrix from the entries of a vector.
In addition, post-multiplication by $U_{\vt q}$ will introduce a product $\tr{\mt B}(\vt q) U_{\vt q}$, which is equal to $-2 {\vt \tau}_b$ \cite{Silveira_2017}, where ${\vt \tau}_b$ is the body-fixed frame representation of the torque.
Then, the fact that ${\mt S}(\vt x) {\vt y} = -{\mt S}(\vt y) {\vt x}$ leads to
\begin{equation*}
\tr{K_{\vt q}} K_{\vt \pi \vt \pi} U_{\vt q} = \frac{1}{2} \tr{\vt \omega} {\mt S}({\mt I}^{-1} {\vt \tau}_b) {\mt I} {\vt \omega}.
\end{equation*}

The term above is identically null, as it corresponds to the quadratic form built with a skew-symmetric matrix.
Another helpful identity taken from Appendix B of Ref.~\citenum{Silveira_2017} is ${\mt B}(\vt q){\vt \omega} = ( \sum_{k=1}^3 \omega_k \hat{\mt B}_k ) \vt q$, where each $\hat{\mt B}_k$ is a skew-symmetric permutation matrix (i.e. $\tr{\hat{\mt B}}_k = -\hat{\mt B}_k$).
This allows us to obtain
\begin{align*}
K_{\vt \pi \vt q} &= \frac{1}{2} \left(\diff{\tr{\vt \omega}}{\vt q}\right) \tr{\mt B}(\vt q) + \frac{1}{2} \sum_{k=1}^3 \omega_k \tr{\hat{\mt B}}_k = \\
&= -\frac{1}{4} {\mt B}(\vt \pi) {\mt I}^{-1} \tr{\mt B}(\vt q) - \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k.
\end{align*}

Pre-multiplication by $\tr{K_{\vt \pi}}$ and post-multiplication by $U_{\vt q}$, followed by some algebraic transformations, ultimately lead to
\begin{equation*}
\tr{K_{\vt \pi}} K_{\vt \pi \vt q} U_{\vt q} = -\frac{1}{2} \tr{\vt \omega} \tr{\mt S}({\mt I}^{-1} {\vt \tau}_b) {\mt I} {\vt \omega} + \frac{1}{2} \tr{\vt \omega} {\mt S}({\vt \tau}_b){\vt \omega}.
\end{equation*}

For the same reason explained above, both terms in the right-hand side of the equation above are identically null.
We are now able to evaluate $\Liu A \Liu A \Ham B $, which is
\begin{align*}
\Liu A \Liu A \Ham B = &\tr{\vt p} {\mt M}^{-1} U_{\vt r \vt r} {\mt M}^{-1} {\vt p} + \tr{\vt \omega} \tr{\mt B}(\vt q) U_{\vt r \vt q} {\mt M}^{-1} {\vt p}
+ \\
&\frac{1}{4} \tr{\vt \omega} \tr{\mt B}(\vt q) U_{\vt q \vt q} {\mt B}(\vt q) \vt \omega.
\end{align*}

Finally, the term that remains for obtaining $\Ham{S}$ is given by
\begin{align*}
\Liu B \Liu A \Ham B &= -\tr{U_{\vt r}} \tr{K_{\vt p \vt p}} U_{\vt r} - \tr{U_{\vt q}} \tr{K_{\vt \pi \vt \pi}} U_{\vt q} = \\
&= -\tr{\vt F} {\mt M}^{-1} {\vt F} - \tr{\vt \tau}_b {\mt I}^{-1} {\vt \tau}_b.
\end{align*}

It is convenient to rewrite the total kinetic energy of the original system, Eq.~\eqref{eq:kinetic energy}, as a quadratic form $K = \frac{1}{2}\tr{[\substack{\vt p \\ \vt \pi}]} {\mt \Omega}(\vt q) [\substack{\vt p \\ \vt \pi}]$, where $\mt \Omega$ is a symmetric, block-diagonal matrix defined as
\begin{equation}
\label{eq:block-diagonal inverse mass tensor}
{\mt \Omega} = \left[\begin{array}{cc}
{\mt M}^{-1} & \mt 0 \\
\mt 0 & \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q)
\end{array}\right].
\end{equation}

We now proceed to present the final expression of $\Ham{S}$, which results from grouping the terms corresponding to forces and torques in the refined potential energy $\refined U(\vt r, \vt q, h^2)$, given by
\begin{equation*}
\refined U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A} {\mt I}^{-1} {\mt A} {\vt \tau} \right),
\end{equation*}
while the terms that depend on the momentum vectors $\vt p$ and $\vt q$ are grouped together in the refined kinetic energy $\refined K$, which can be written in a compact matrix form as
\begin{equation*}
\refined K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \refined{\mathbf \Omega}(\vt r, \vt q, h^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
\end{equation*}
where
\begin{equation*}
\refined{\mt \Omega} = {\mt \Omega} + \frac{\timestep^2}{6} {\mt \Omega} \left[\begin{array}{cc}
U_{\vt r \vt r} & U_{\vt r \vt q} \\
U_{\vt q \vt r} & U_{\vt q \vt q}
\end{array}\right] \tr{\mt \Omega}.
\end{equation*}

Finally, the shadow Hamiltonian is given by $\Ham{S} = {\refined U} + {\refined K} + {\mathcal O}(\timestep^4)$.
